Dotted Suffix Trees\\A Structure for Approximate Text Indexing

### 2005-09-16 paper/abstract.tex
Many small changes and it's almost done

The problem we address is text indexing for approximate matching. We consider that we are given a text $\mathcal{T}$ which undergoes some preprocessing to generate an index. We can later query this index to identify the places where a string occurs up to a certain number of allowed substitutions $k$. We present a structure for indexing which occupies space $\mathcal{O}(n\log^kn)$ in the average case, independent of alphabet size, $n$ being the text size. This structure support searching in $\mathcal{O}(m^{k+1})$ time, for patterns of size $m$, again independent of alphabet size. The construction of the structure has time bound by $\mathcal{O}(N|\Sigma|)$, where $N$ is the number of nodes in the index and $|\Sigma|$ the alphabet size.

>>>>

The problem we address is text indexing for approximate matching. We consider that we are given a text $\mathcal{T}$ which undergoes some preprocessing to generate an index. We can later query this index to identify the places where a string occurs up to a certain number of allowed substitutions $k$. We present a structure for indexing which occupies space $\mathcal{O}(n\log^kn)$ in the average case, independent of alphabet size, $n$ being the text size. This structure support searching in $\mathcal{O}(m^{k+1})$ time, for patterns of size $m$, again independent of alphabet size. The construction of the structure has time bound by $\mathcal{O}(kN|\Sigma|)$, where $N$ is the number of nodes in the index and $|\Sigma|$ the alphabet size.

### 2005-09-16 paper/conclusion.tex
Many small changes and it's almost done

We have presented an indexing structure for approximate text matching which takes $\mathcal{O}(n\log^k n)$ space in the average case (this can be seen either in a theoretical or experimentally). This structure allows matches in $\mathcal{O}(m^{k+1})$ time and can be constructed in $\mathcal{O}(N|\Sigma|)$ time ($N$ being the actual number of nodes). The structure and the algorithms to construct it are very simple and should be easily implemented.

@@

\subsection{Acknowledgements}

We thank Y for making their suffix tree implementation available~\cite{} over which we implemented our algorithms.

>>>>

We have presented an indexing structure for approximate text matching which takes $\mathcal{O}(n\log^k n)$ space in the average case (this can be seen either in a theoretical or experimentally). This structure allows matches in $\mathcal{O}(m^{k+1})$ time and can be constructed in $\mathcal{O}(kN|\Sigma|)$ time ($N$ being the actual number of nodes). The structure and the algorithms to construct it are very simple and should be easily implemented.

@@

\subsection{Acknowledgments}

We thank Lu\'{\i}s~Russo and Sara~Madeira for several productive discussions.

### 2005-09-16 paper/construction.tex
Many small changes and it's almost done

The construction of the tree using either Ukonnen or McCreight's algorithm assures one that this operation is always correct due to the way that the indices are created.

@@

\subsection{Extension to any number of errors}

@@

We note that the method of removing the existing error trees can be improved in practice (the asymptotic bound remains the same) the following way: While making a copy of the tree for construction of the root's error tree, instead of following a dot link to make a copy, move the error tree to its destination and make the necessary adjustments inplace (this move will probably be just an adjustment of pointer depending on the exact implementation). Using this method only leafs labeled one need to be removed (as well as any internal node which becomes unnecessary).

The analysis for the time cost above remains valid. Taking $N_i$ to mean the number of nodes of the i-error tree, we now have that the time cost is $\mathcal{O}(N_1|\Sigma|+N_2|\Sigma|+\dots+N_k|\Sigma)=\mathcal{O}((N_1+N_2+\dots+N_k)|\Sigma|)=\mathcal{O}(N_k|\Sigma|)=\mathcal{O}(N|\Sigma|)$ since $N_k \le N_i$ ($N_k = N$).

>>>>

The construction of the tree using either Ukonnen's or McCreight's algorithm assures one that this operation is always correct due to the way that the indices are created.

@@

\subsection{Extension to any number of errors}\label{subsec:construct-k-larger-1}

@@

We note that the method of removing the existing error trees can be improved in practice (the asymptotic bound remains the same) the following way: While making a copy of the tree for construction of the root's error tree, instead of following a dot link to make a copy, move the error tree to its destination and make the necessary adjustments in place (this move will probably be just an adjustment of pointer depending on the exact implementation). Using this method only leafs labeled one need to be removed (as well as any internal node which becomes unnecessary).

\begin{definition}
We define $N_k$ to be the number of nodes of the k-error dotted tree.

$N_0$ is the number of nodes of the suffix tree.

For a known $k$, we set $N=N_k$.
\end{definition}

The analysis for the time cost above remains valid. We now have that the time cost is $\mathcal{O}(N_1|\Sigma|+N_2|\Sigma|+\dots+N_k|\Sigma)=\mathcal{O}((N_1+N_2+\dots+N_k)|\Sigma|)=\mathcal{O}(kN|\Sigma|)$.

### 2005-09-16 paper/experimental.tex
Many small changes and it's almost done

The code was implemented in the C~Language, starting from the suffix tree implementation by FIXME.

@@

\begin{enumerate}
\item ``Democracy in America'' by Alexis de Tocqueville as natural language example
\item The DNA of S. Cervesiae (first chromossome)
\item Randomly generated text
\end{enumerate}

@@

\begin{figure}%
\includegraphics{ratio-1}%
\includegraphics{ratio-2}%
\includegraphics{ratio-3}%
\caption{Ratios of k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure}

To experimentally verify the average case, we show in Figure~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees. We can easily see that for all three data sets, the experimental values do resemble a logarithm.

@@

For searching, the full textes were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen bits of the text were taken. These patterns were then altered by substituting a character at random. To generate mismatches, one character too many was altered~\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurences (ie, it only reported whether the string exists in the text). Therefore, the number of occurences had no influence on the search time.

>>>>

The code was implemented in the C~Language, starting from an open-source implementation~\cite{ansicimplementation}.

@@

\begin{enumerate}
\item The English translation of ``Democracy in America'' by Alexis de Tocqueville as natural language example
\item The DNA of S. Cervesiae (first chromosome)
\item Randomly generated text by a uniform distribution
\end{enumerate}

@@

\begin{figure}%
\centering
\subfigure[English Text]{
\includegraphics[width=.25\textwidth]{ratio-in=english-0-1.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=english-1-2.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=english-2-3.eps}%
}
\\
\subfigure[DNA]{
\includegraphics[width=.25\textwidth]{ratio-in=dna-0-1.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=dna-0-1.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=dna-1-2.eps}%
}
\\
\subfigure[Random Text]{
\includegraphics[width=.25\textwidth]{ratio-in=random-0-1.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=random-0-1.eps}%
\includegraphics[width=.25\textwidth]{ratio-in=random-1-2.eps}%
}
\caption{Ratios of k-error and (k+1)-error dotted trees}\label{fig:ratios}%
\end{figure}

To experimentally verify the average case, we show in Figures~\ref{fig:ratios} the ratios between the k-error and the (k+1)-error dotted trees. We can easily see that for all three data sets, the experimental values do resemble a logarithm as predicted.

@@

For searching, the full texts were first indexed and then online searches were performed on top of the structure. To generate input patterns, randomly chosen bits of the text were taken. These patterns were then altered by substituting a character at random. To generate mismatches, one character too many was altered~\footnote{Since the characters used were not present in the original text, there is no risk of generating a string which \emph{does appear} in the text.}. The search algorithm performed an early exit, without reporting of occurrences (ie, it only reported whether the string exists in the text). Therefore, the number of occurrences had no influence on the search time.

\begin{figure}
\includegraphics[width=\textwidth]{search-time-vary-m.k=2.in=english-100k.eps}
\caption{Search time versus pattern size}\label{fig:search-vary-m}
\end{figure}

Figure~\ref{fig:search-vary-m} shows the search time for growing pattern lenghts. It shows the case where the pattern exists and where it does not (an extra error is added). The fact that the search is done using an early-exit strategy explains why non existing patterns take longer than existing patterns. The value presented is the number of character comparisons the algorithm makes.

### 2005-09-16 paper/search.tex
Many small changes and it's almost done

To understand the pseudo-code, it is important to understand that suffix trees are normally implemented as shown in Figure~\ref{fig:mississippi-nodes} where each node contains the start and end indices of its incoming edge. When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$.

@@

This algorithm runs in time $O(m^{k+1})$:

The recursive call at line~\ref{algoline:search:followDot} means that a dot link was followed. There are at most $\mathcal{O}(m^k)$ places where a dot link can be followed while descending a string~\footnote{For $k=1$ each point in the string. For $k>1$, each path for $k-1$, can now have at most $m$ extra branches.}. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurences (which can be done in time linear to the number of occurences).

>>>>

To understand the pseudo-code, it is important to understand that suffix trees are normally implemented so that each node tracks the string on the edge leading to it by keeping the start and end indices on the original string (all strings on the trees presented here are substrings of the original one). On Figure~\ref{fig:mississippi-nodes} we show an example of this structure (for the now familiar \putstring{mississippi}) where the string each node references is shown explicitly.

When presenting the algorithm, we take the liberty of using $s+j$ where $s$ is a string and $j$ an integer to mean the suffix of $s$ starting at position $j$ as happens in the C~Programming Language.

@@

This algorithm runs in time $\mathcal{O}(m^{k+1})$:

The recursive call at line~\ref{algoline:search:followDot} means that a dot link was followed. There are at most $\mathcal{O}(m^k)$ places where a dot link can be followed while descending a string~\footnote{For $k=1$ each point in the string. For $k>1$, each path for $k-1$, can now have at most $m$ extra branches.}. Following one path in the tree takes $\mathcal{O}(m)$ time. Therefore, the total time is $\mathcal{O}(m^{k+1})$ without the reporting of occurrences (which can be done in time linear to the number of occurrences).

### 2005-09-16 paper/space.tex
Make the aaaa figure smaller
Many small changes and it's almost done

\begin{figure}
\includegraphics{figures/aaaa}
\caption{$a^n$: worst case space consumption}%
\label{fig:aaaa}
\end{figure}

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = aaa\cdots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $\mathcal{O}(n^{k+1})$ nodes.

@@

\begin{definition}
We define $N_k$ to be the number of nodes of the k-error dotted tree.

$N_0$ is the number of nodes of the suffix tree.
\end{definition}

@@

\[ l = \max_{\overline{w} \in \mathcal{T}} |w| \mathrm{and \overline{w} isn't a leaf} \]
\end{definition}

@@

So far, we have achieved little since in the worst case $l=n-1$ as shown above. However, the expected case is $l=\mathcal{O}(\log n)$~\cite{apostolico92selfalignments} and we have $N_k=\mathcal{O}(n\log^k n)$.

>>>>

\begin{figure}
\centering
\includegraphics[height=.2\textheight]{figures/aaaa}
\caption{$a^n$: worst case space consumption}%
\label{fig:aaaa}
\end{figure}

How much space does a k-error dotted tree take? Suffix Trees have a number of nodes proportional to the size of the string (which we will refer to as $n$). One-error dotted trees have more nodes. The worst case presented in Figure~\ref{fig:aaaa} is a string of the form $a^n = aaa\dots$ which generates a one-dotted tree taking $\mathcal{O}(n^2)$ nodes. For a larger number of errors $k$, it is easy to see that this leads to $\mathcal{O}(n^{k+1})$ nodes.

@@

@@

\[ l = \max_{\overline{w} \in \mathcal{T}} |w| \mbox{ and $\overline{w}$ isn't a leaf} \]
\end{definition}

@@

So far, we have achieved little since in the worst case $l=n-1$ as shown above. However, the expected case is $l=\mathcal{O}(\log n)$~\cite{apostolico92selfalignments,szpankowski:unexpected} and we have $N_k=\mathcal{O}(n\log^k n)$.
